---
title: "Define Custom Loss Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Define Custom Loss Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(threedx)
```

Since `learn_weights()` fits a `threedx` model by evaluating all parameter combinations provided via the `alphas_grid` argument, we can use any kind of `loss_function` to compare models on the training data.

While `threedx` comes with a few built-in loss functions, they serve as defaults and as examples. You can define any other that fits your application, and you're not limited to those for which you know the derivative.

## Constructing a Loss Function

At the time of writing, `learn_weights()` passes three arguments to a loss function:

- A numeric vector of observed actuals `y` without missing values
- A numeric vector of one-step-ahead predictions `y_hat` of same length as `y` and also without missing values
- Further arguments `...` that might be provided in the future

For a loss function, the access to `y` and `y_hat` of course is essential to compute errors, or to assign weights to errors.

In return, `learn_weights()` expects to receive a non-missing numeric scalar value from the loss function. These scalars will be minimized across parameter combinations defined by `alphas_grid` to choose the optimal parameter combination.

One of the simplest possible loss functions that can be used is the mean absolute error loss function. Its implementation in `threedx` is straightforward:

```{r}
print(loss_mae)
```

```{r}
loss_mae(y_hat = c(5, -3, 2), y = c(1, 1, -2))
```

But you can go more complex. For example, `threedx` provides a loss that assigns weight (based on the size of the observation) to each absolute error before taking the mean:

```{r}
print(loss_mae_with_observation_weight)
```

Wether this is a useful loss function will depend on your context. But it goes to show that you have a lot of flexibility.

Another idea could be to define a function that only ever takes the 20 most recent observations into account:

```{r}
loss_mae_twenty_most_recent <- function(y_hat, y, ...) {
  loss_mae(utils::tail(y_hat, 20), utils::tail(y, 20), ...)
}
```

Or one that applies moving averages on observations and predictions first, or one that only takes into account the observations from the same seasonal period as the next one-step-ahead prediction will perform, or ...