---
title: "Working with Forecast Sample Paths"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Forecast Sample Paths}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
set.seed(636)
library(threedx)
```

Forecast sample paths are the best. In contrast to quantiles, you can aggregate them!

Most forecast model implementations only aim to provide select quantiles of the forecast distribution to quantify the forecast uncertainty---if at all. While this suffices for applications that are strictly defined and will not extend their use case in the future, quantiles can be very restrictive in a business context where one needs to perform arbitrary and often unforeseen aggregations across time horizons, hierarchies, and the like.

If you only have quantiles, you can't do it. Quantiles can't be aggregated.

Maybe you can rely on an assumption of independent Normal distributions, and sum those. But auto-correlation would get lost.

In contrast, sample paths are neat because they provide a representation of the joint forecast distribution over time horizons. In contrast to marginal quantiles, you can first aggregate sample paths in arbitrary ways, and *then take quantiles* to summarize the distribution of an arbitrary aggregation of interest.

Let's fit a model...

```{r}
y <- rpois(n = 55, lambda = pmax(0.1, 1 + 10 * sinpi(6:59 / 6)))

model <- threedx::learn_weights(
  y = y,
  period_length = 12L,
  alphas_grid = threedx::list_sampled_alphas(
    n_target = 1000L,
    include_edge_cases = TRUE
  ),
  loss_function = loss_mae
)
```

... and generate forecast sample paths:

```{r}
forecast <- predict(
  object = model,
  horizon = 12L,
  n_samples = 2500L,
  observation_driven = TRUE
)
```

The forecasts are stored in a `threedx_paths` object,

```{r}
class(forecast)
```

of which the main component is a matrix of sample paths (each row is a sample path, each column a forecast horizon):

```{r}
# printing only the first 5 of `n_samples` sample paths
forecast$paths[1:5, ]
```

### Summarize Over Sample Paths

To summarize the forecast at each horizon, we can compute the mean...

```{r}
round(colMeans(forecast$paths), 1)
```

... or any quantile...

```{r}
round(apply(forecast$paths, 2, quantile, 0.9), 1)
```

... or ask specific questions such as "what is the probability of an observation larger than zero?":

```{r}
round(apply(forecast$paths, 2, function(x) mean(x > 0)), 2)
```

### Summarize Over Horizons

Similarly, we can summarize along the `n_samples` dimension of the matrix.

We can ask for the total value across all horizons by sample path...

```{r}
rowSums(forecast$paths)[1:5]
```

... or for the total value in the six-month period after horizon three...

```{r}
rowSums(
  forecast$paths[, 4:9]
)[1:5]
```

... or how many of the next twelve months will exhibit a value of zero:

```{r}
rowSums(
  forecast$paths == 0
)[1:5]
```

### Summarize Over Horizons And Sample Paths

Usually, we want to follow up an aggregation across horizons by an aggregation across sample paths.

We can get the median sum for the next twelve months...

```{r}
median(rowSums(forecast$paths))
```

... the 95% quantile for the six-month period after horizon three...

```{r}
quantile(rowSums(forecast$paths[, 4:9]), 0.95)
```

... or the probability of zero value at horizons 7 through 9:

```{r}
mean(rowSums(forecast$paths[, 7:9]) == 0)
```

## Visualize Sample Path Forecasts

If you have `ggplot2` installed, `threedx` provides two ways of visualizing a `threedx_paths` object using the `autoplot()` method.

```{r}
library(ggplot2)
```

The default is a graph of marginal quantiles of the sample paths for each forecast horizon as many other forecast packages would provide it:

```{r}
autoplot(forecast)
```

An alternative option specified via the argument `method = "paths"` is the visualization of individual sample paths as lines. Since a graph of sample paths becomes very crowded very quickly, by default only `n = 5` randomly chosen paths are plotted:

```{r}
autoplot(forecast, method = "paths")
```

While forecast quantiles sometimes give the impression that a model expects a flat line of future observations. Visualizations of sample paths make clear that any given future scenario is anything but a flat line.
