---
title: "Forecasting Trends with `threedx`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forecasting Trends with `threedx`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
set.seed(637)
library(threedx)
library(ggplot2)
```

The `threedx` model is not designed to forecast trends in time series. As its predictions are a pure weighted average of historical observations, the model does not extrapolate beyond the limits of the historical data. Nevertheless, one can choose the **innovation function** used during generation of the forecast paths to capture the systematic error of the model.

To demonstrate the issue, let's start with a simple trending time series that clearly continues to trend upward and has fairly little noise, and no trend component.

```{r example_trend}
df <- data.frame(
  date = seq(as.Date("2019-01-01"), as.Date("2023-12-01"), by = "month"),
  y = sqrt(1:60) * 5 + rnorm(60)
)

df_train <- df[df$date < as.Date("2023-01-01"), ]
df_test <- df[df$date >= as.Date("2023-01-01"), ]
```

```{r plot_example_trend, echo = FALSE}
ggplot(df, aes(x = date, y = y)) +
  geom_line(color = "grey") +
  geom_point() +
  geom_vline(aes(xintercept = as.Date("2022-12-15")), linetype = 3)
```

The best `threedx` can do, with no trend component to speak of, is to base the next observation to nearly the full extent on the most recent observation. When one can't predict the trend, predicting the same level from the previous observation is the best one can do. The forecast is a straight line, while the actuals continue to rise:

```{r forecast_trend}
model <- learn_weights(
  y = df_train$y,
  period_length = 12L,
  alphas_grid = list_sampled_alphas(include_edge_cases = TRUE),
  loss_function = loss_mae
)

predict(
  object = model,
  horizon = 12L,
  n_samples = 1000L,
  observation_driven = FALSE,
  innovation_function = draw_normal_with_zero_mean
) |>
  autoplot(date = df_train$date, date_future = df_test$date) +
  geom_line(aes(y = y), data = df_test, color = "grey") +
  geom_point(aes(y = y), data = df_test) +
  geom_vline(aes(xintercept = as.Date("2022-12-15")), linetype = 3)
```

If we inspect the model, we see that the one-step-ahead residuals have a clear bias (positive mean and median) as the model doesn't predict the on-average increase month-over-month:

```{r model_residuals}
print(summary(model$residuals))
```

We can make use of this bias in the residuals at prediction time.

By choosing an innovation function that does not enforce a zero-mean assumption on the innovation samples projected into the future (as standard implementations of forecasting approaches such as ARIMA and ETS do), we can add this bias onto the forecast, and thereby offset the model's bias.

Using `draw_normal_with_drift()` will use the sample mean of the residuals to draw innovations (in contrast to the hardcoded mean at zero used in `draw_normal_with_zero_mean()` above):

```{r forecast_with_drift}
predict(
  object = model,
  horizon = 12L,
  n_samples = 1000L,
  observation_driven = FALSE,
  innovation_function = draw_normal_with_drift
) |>
  autoplot(date = df_train$date, date_future = df_test$date) +
  geom_line(aes(y = y), data = df_test, color = "grey") +
  geom_point(aes(y = y), data = df_test) +
  geom_vline(aes(xintercept = as.Date("2022-12-15")), linetype = 3)
```

We can also drop the normality assumption and bootstrap entirely from residuals (which here doesn't make a big difference as the time series was generated from a Normal distribution):

```{r forecast_with_bootstrap}
predict(
  object = model,
  horizon = 12L,
  n_samples = 1000L,
  observation_driven = FALSE,
  innovation_function = draw_bootstrap
) |>
  autoplot(date = df_train$date, date_future = df_test$date) +
  geom_line(aes(y = y), data = df_test, color = "grey") +
  geom_point(aes(y = y), data = df_test) +
  geom_vline(aes(xintercept = as.Date("2022-12-15")), linetype = 3)
```

## The Innovation Function Is Not A Panacea

While the choice of the innovation function is a remedy for the missing trend component in the above example, it doesn't work quite as neatly in all cases.

By using an innovation function that doesn't enforce a zero-mean, the forecast can be improved for time series that consist of a strong and mostly linear trend.

However, if the trend is not as uniform across past observations, or if there are seasonal components as well, the forecast can be much trickier. 

As soon as the model tries to compensate the trend in some way during training (instead of using an `alpha` close to 1), and thereby messes up a clean prediction of the seasonal component, there is not much that can be fixed by the innovation function choice.

Consider the following extreme example to make the problem tangible. We will forecast a monthly seasonal time series that trends upward with every year by three additional units. To make the relationship between trend, model, and residuals clear, we don't add any noise component:

```{r example_trend_and_season}
component_season <- rep(c(0,1,2,3,4,5,6,5,4,3,2,1), times = 5) * 5
component_trend <- 1:60

y <- component_season + component_trend
```

```{r plot_example_trend_and_season, echo = FALSE}
ggplot(data.frame(y = y, index = seq_along(y)), aes(x = index, y = y)) +
  geom_line(color = "grey") +
  geom_point()
```

First, note that `threedx` can forecast the seasonal component perfectly if it's the *only* component:

```{r forecast_component_season}
threedx::learn_weights(
  y = component_season,
  period_length = 12L,
  alphas_grid = list_sampled_alphas(include_edge_cases = TRUE),
  loss_function = loss_mae
) |>
  predict(
    horizon = 24L,
    n_samples = 1000L,
    observation_driven = FALSE,
    innovation_function = draw_normal_with_zero_mean
  ) |>
  autoplot()
```

However, when trend and seasonality are combined, the model training results in an unfortunate parameter choice that relinquishes the perfect seasonality prediction to accompany the trend in a desperate fashion.

The reason for this is the loss function: The mean absolute error (used here via `loss_mae()`) is reduced by trading off the error due to the trend with the error due to the seasonality. It does not accept the bias in the one-step-ahead predictions that would remain due to the trend for models that predict the seasonality perfectly.

```{r forecast_trend_and_season}
threedx::learn_weights(
  y = y,
  period_length = 12L,
  alphas_grid = list_sampled_alphas(include_edge_cases = TRUE),
  loss_function = loss_mae
) |>
  predict(
    horizon = 12L,
    n_samples = 1000L,
    observation_driven = FALSE,
    innovation_function = draw_normal_with_drift
  ) |>
  autoplot()
```

We can recover the optimal model choice (in combination with the drifting innovation function) by using a loss function that ignores the bias. Let's redefine the mean absolute error, removing the mean bias:

```{r loss_mae_ignoring_bias}
loss_mae_ignoring_bias <- function(y_hat, y, ...) {
  residuals <- y - y_hat
  residuals_zero_bias <- residuals - mean(residuals)
  mean(abs(residuals_zero_bias))
}
```

Using `loss_mae_ignoring_bias()` instead of `loss_mae()`, the model that perfectly removes the seasonal error is chosen as the trend bias is ignored during training. During prediction, the bias is compensated by the innovations with drift.

```{r train_with_loss_mae_ignoring_bias}
threedx::learn_weights(
  y = y,
  period_length = 12L,
  alphas_grid = list_sampled_alphas(include_edge_cases = TRUE),
  loss_function = loss_mae_ignoring_bias
) |>
  predict(
    horizon = 12L,
    n_samples = 1000L,
    observation_driven = FALSE,
    innovation_function = draw_normal_with_drift
  ) |>
  autoplot()
```

But again, this is unlikely to be a panacea.
